---
title: "command"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{command}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(command)
library(fs)
```



The only way to write complex software that won't fall on its face is to build it out of simple modules connected by well-defined interfaces, so that most problems are local and you can have some hope of fixing or optimizing a part without breaking the whole.
http://catb.org/%7Eesr/writings/taoup/html/modularitychapter.html



# The problem of writing safe, flexible data analysis code

- classic example: big file
    - difficult to understand
    - slow or confusing/unreliable - have to re-run
    - difficult to debug
- split into logical parts (modular), then source
    - helps
    - but doesn't totally solve - confusing environment
- better: break into pieces, and run each in own environment
- a bit like functions (but without global environment)
- push harder - make inputs outputs completely transparenet

    have to re-run - slow
    - 
    - difficult to change
    - polluted environment
- classic software solution: break into small files
- but if 
- how to run them?

[https://stemurphy.com/post/rep_manu_think_about/]

# Other solutions

- commandArgs
- docopt
- other?




https://swcarpentry.github.io/r-novice-inflammation/06-best-practices-R.html

"Another way you can be explicit about the requirements of your code and improve it’s reproducibility is to limit the “hard-coding” of the input and output files for your script. If your code will read in data from a file, define a variable early in your code that stores the path to that file. For example"




```{r, root_dir}
root_dir <- system.file("extdata/swiss_fertility", 
                        package = "command")
dir_tree(root_dir)
```


```{r, echo = FALSE, comment = ""}
cat(readLines(file.path(root_dir, "run_all.sh")), sep = "\n")
```



```{r, echo = FALSE}
cat(readLines(file.path(root_dir, "src", "dataset.R")), sep = "\n")
```

