---
title: "Creating a Shell Script"
---

```{r, include = FALSE}
library(command)
library(fs)

path_swiss <- path_package(package = "command",
                           "extdata/swiss1")
dir_copy(path = path(path_swiss, "data"),
         new_path = ".")
dir_create("src")	 
file_copy(path = path(path_swiss, "src/cleaned_data.R"),
          new_path = "src/cleaned_data.R")
file_copy(path = path(path_swiss, "src/model.R"),
          new_path = "src/model.R")	
file_copy(path = path(path_swiss, "src/vals_fitted.R"),
          new_path = "src/vals_fitted.R")	
file_copy(path = path(path_swiss, "src/fig_fitted.R"),
          new_path = "src/fig_fitted.R")	
dir_create("out")	 
file_copy(path = path(path_swiss, "report.qmd"),
          new_path = "report.qmd")

knitr::opts_chunk$set(collapse = TRUE,
                      engine.opts = "-l",
                      comment = "")

old_hooks <- fansi::set_knit_hooks(knitr::knit_hooks) ## so colors work

nms_articles <- "a1_quickstart.Rmd|a2_shell_script.Rmd|a3_makefile.Rmd|a4_workflow.Rmd|references.bib|makefile_deps.png"
```

# Introduction

Functions `cmd_shell()` and `shell_script()` help with writing a shell script to control a data analysis workflow. `cmd_shell()` and `shell_script()` do most of the work, but some additional editing is usually necessary.

# An example workflow

To demonstrate, we use a simple workflow where all the code, aside from the shell script, has already been written. The project directory contains the following files:

```{r, echo = FALSE}
dir_tree(regexp = nms_articles,
         invert = TRUE)
```

The workflow has five steps:

1. `src/cleaned_data.R` reads in the raw data, cleans it, and creates `out/cleaned_data.rds`
2. `src/model.R` fits a model to the cleaned data, and creates `out/model.rds`
3. `src/vals_fitted.R` extract fitted values from the model, and creates `vals_fitted.rds`
4. `src/fig_fitted.R` uses `vals_fitted.rds` to create the plot in `out/fig_fitted.png`
5. `report.qmd` uses `out/fig_fitted.png` to create the document `report.html`

`cleaned_data.R`, `model.R`, and `vals_fitted.R`, and `fig_fitted.R` all contain calls to `cmd_assign()`. For instance, `src/model.R` contains the lines

```{r, echo = FALSE}
lines <- readLines("src/model.R")
cat(paste(c(lines[7:8], ""), collapse = "\n"))
```

# cmd_shell()

`cmd_shell()` extracts the call to `cmd_assign()` from an R file, and turns it into a shell command. For instance, the call

```{r, eval = FALSE}
cmd_shell("src/model.R")
```
has return value
```{r, eval = TRUE, echo = FALSE}
cmd_shell("src/model.R")
```

The easiest way to use `cmd_shell()` is to call it from the R console, and then cut-and-paste the results into the shell script.


# shell_script()

`shell_script()` creates a draft of the whole shell script. `shell_script()` loops through all the R files in a directory, extracting `cmd_assign()` calls and converting them into shell commands. It then puts the shell commands into a script.

For instance, the call
<!-- Messages don't display properly, so don't print real output -->
```{r, eval = FALSE}
shell_script("src")
```
```{r, include = FALSE}
shell_script("src")
```
creates a new file called `workflow.sh` with the lines
```{r, echo = FALSE}
lines <- readLines("workflow.sh")
cat(paste(lines, collapse = "\n"))
```

Here, as in most cases, the output from `shell_script()` needs some editing before it is ready for use. The ordering of the commands in `workflow.sh` reflects the ordering of the files in the `src` directory. We need to rearrange the commands to instead reflect the order in which they need to be executed. We also add instructions on rendering the file `report.qmd`.

```{r, echo = FALSE}
lines <- lines[c(2:5, 10:17, 1, 6:9)]
lines <- c(lines,
	   "quarto render report.qmd",
	   "")
cat(paste(lines, collapse = "\n"))
writeLines(lines, con = "workflow.sh")
```

# Running the shell script

Our project directory looks like this.

```{r, echo = FALSE}
dir_tree(regexp = nms_articles,
         invert = TRUE,
	 recurse = 1L)
```

We run the shell script.

```{bash}
bash workflow.sh
```

Our project directory now looks like this.

```{r, echo = FALSE}
dir_tree(regexp = nms_articles,
         invert = TRUE,
	 recurse = 1L)
```

We have created `cleaned_data.rds`, `model.rds`, and `vals_fitted.rds`, and `fig_fitted.png` in the `out` directory, and `report.html` in the main directory. The `report_files` directory was created by quarto.

# Other resources

- [makefile()](https://bayesiandemography.github.io/command/reference/makefile.html) The Makefile equivalent of `shell_script()`
- [Data Analysis Workflows](https://bayesiandemography.github.io/command/articles/workflow.html)
  Safe, flexible workflows for data analysis
- Episodes 1--3 of [The Unix Shell](https://swcarpentry.github.io/shell-novice/index.html)
  Introduction to the command line
- [Command-Line Programs](https://swcarpentry.github.io/r-novice-inflammation/05-cmdline.html)
  Introduction to Rscript



```{r, include = FALSE}
files_to_delete <- dir_ls(regexp = nms_articles, 
                          invert = TRUE)
file_delete(files_to_delete)
```



