---
title: "Creating a Makefile"
---

```{r, include = FALSE}
library(command)
library(fs)

path_swiss <- path_package(package = "command",
                           "extdata/swiss1")
dir_copy(path = path(path_swiss, "data"),
         new_path = ".")
dir_create("src")	 
file_copy(path = path(path_swiss, "src/cleaned_data.R"),
          new_path = "src/cleaned_data.R")
file_copy(path = path(path_swiss, "src/model.R"),
          new_path = "src/model.R")	
file_copy(path = path(path_swiss, "src/vals_fitted.R"),
          new_path = "src/vals_fitted.R")	
file_copy(path = path(path_swiss, "src/fig_fitted.R"),
          new_path = "src/fig_fitted.R")	
dir_create("out")	 
file_copy(path = path(path_swiss, "report.qmd"),
          new_path = "report.qmd")

knitr::opts_chunk$set(collapse = TRUE,
                      engine.opts = "-l",
                      comment = "")

old_hooks <- fansi::set_knit_hooks(knitr::knit_hooks) ## so colors work

nms_articles <- "a1_quickstart.Rmd|a2_shell_script.Rmd|a3_makefile.Rmd|a4_workflow.Rmd|references.bib|makefile_deps.png"
```


# Introduction

Functions `cmd_make()` and `makefile()` help with writing a Makefile to control a data analysis workflow. `cmd_make()` and `makefile()` do much of the work, but some extra hand coding is also necessary.

# An example workflow

To demonstrate, we use a simple workflow where all the code, aside from the shell script, has already been written. The project directory contains the following files:

```{r, echo = FALSE}
dir_tree(regexp = nms_articles,
         invert = TRUE)
```

1. `src/cleaned_data.R` reads in the raw data, cleans it, and creates `out/cleaned_data.rds`
2. `src/model.R` fits a model to the cleaned data, and creates `out/model.rds`
3. `src/vals_fitted.R` extract fitted values from the model, and creates `vals_fitted.rds`
4. `src/fig_fitted.R` uses `vals_fitted.rds` to create the plot in `out/fig_fitted.png`
5. `report.qmd` uses `out/fig_fitted.png` to create the document `report.html`

`cleaned_data.R`, `model.R`, and `vals_fitted.R`, and `fig_fitted.R` all contain calls to `cmd_assign()`. For instance, `src/model.R` contains the lines

`cleaned_data.R`, `model.R`, and `vals_fitted.R`, and `fig_fitted.R` all contain calls to `cmd_assign()`. For instance, `src/model.R` contains the lines

```{r, echo = FALSE}
lines <- readLines("src/model.R")
cat(paste(c(lines[7:8], ""), collapse = "\n"))
```

# cmd_make()

`cmd_make()` extracts the call to `cmd_assign()` from an R file, and turns it into a Makefile rule. For instance, the call

```{r, eval = FALSE}
cmd_make("src/model.R")
```
has return value
```{r, eval = TRUE, echo = FALSE}
cmd_make("src/model.R")
```

The easiest way to use `cmd_make()` is to call it from the R console, and then cut-and-paste the results into Makefile.


# makefile()

`makefile()` creates a draft of the whole Makefile. `makefile()` loops through all the R files in a directory, extracting `cmd_assign()` calls and converting them into Makefile rules. It then puts the rules into a Makefile.

For instance, the call
<!-- Messages don't display properly, so don't print real output -->
```{r, eval = FALSE}
makefile("src")
```
```{r, include = FALSE}
makefile("src")
```
creates a new file called `Makefile` with the lines
```{r, echo = FALSE}
lines <- readLines("Makefile")
cat(paste(lines, collapse = "\n"))
```

The output from `makefile()` needs some editing before it is ready for use. We need to a rule for creating the report, and make that report a prerequisite for `all:` at the top of the file:

```{r, echo = FALSE}
lines[[3L]] <- "all: report.html"
lines <- c(lines[1:22],
           "report.html: report.qmd \\",
	   "  out/tab_coef.rds",
	   "\tquarto render $<",
	   "",
	   lines[-(1:22)])
cat(paste(lines, collapse = "\n"))
```

The ordering of the rules in the Makefile reflects the ordering of the files in the `src` directory, not the order in which the rules should be executed. This does not matter to `make`, which constructs its own dependency map. But for the benefit of human readers, we rearrange the rules to match the execution order.


```{r, echo = FALSE}
lines <- lines[c(2:9, 14:22, 10:13, 1, 23:31)]
cat(paste(lines, collapse = "\n"))
writeLines(lines, con = "Makefile")
```


# Running the Makefile

Our project directory looks like this.

```{r, echo = FALSE}
dir_tree(regexp = nms_articles,
         invert = TRUE,
	 recurse = 1L)
```

We run the Makefile.

```{bash}
#make
```

Our project directory now looks like this.

```{r, echo = FALSE}
dir_tree(regexp = nms_articles,
         invert = TRUE,
	 recurse = 1L)
```

We have created `scaled.rds`, `model.rds`, and `tab_coef` in the `out` directory, and `report.html` in the main directory.


# Other resources

- [shell_script()](https://bayesiandemography.github.io/command/reference/makefile.html) The shell script equivalent of `makefile()`
- [Data Analysis Workflows](https://bayesiandemography.github.io/command/articles/workflow.html)
  Safe, flexible workflows for data analysis
- [Project Management with Make](https://jeroenjanssens.com/dsatcl/chapter-6-project-management-with-make)
  Makefiles in data analysis workflows
- [GNU make](https://www.gnu.org/software/make/manual/make.html#SEC_Contents)
  Definitive guide
- [Command-Line Programs](https://swcarpentry.github.io/r-novice-inflammation/05-cmdline.html)
  Introduction to Rscript


```{r, include = FALSE}
files_to_delete <- dir_ls(regexp = nms_articles, 
                          invert = TRUE)
file_delete(files_to_delete)
```



