---
title: "Using command with renv"
---

```{r, include = FALSE}
library(command)

knitr::opts_chunk$set(collapse = TRUE,
                      comment = "")
```

# Key message

If a project uses package `renv`, all R scripts should start with the line
```{r, eval = FALSE}
command::use_renv()
```


# Introduction

One obstacle to creating reproducible workflows in R is that the behavior of R packages can differ across versions. An R script that behaves one way if it is run using version 1.0 of a package may behave a different way if it is run using version 2.0 of the package. The [renv](https://rstudio.github.io/renv/) package solves this problem by allowing users to lock down the exact versions of the packages a project uses.

The mechanism that `renv` uses to lock packages down is, however, bypassed by [Rscript](https://stat.ethz.ch/R-manual/R-devel/library/utils/html/Rscript.html) or [littler](https://CRAN.R-project.org/package=littler). A workflow that uses Rscript or littler needs to be tweaked before `renv` performs correctly. Function `command::use_renv()` helps with this tweaking.


# How renv can fail with Rscript and littler

A project that uses `renv` contains a script `renv/activate.R`. That script tells R which package versions to use when running project code. `renv` only performs correctly if R runs `renv/activate.R` when the session starts.

`renv` arranges for `renv/activate.R` to be run at startup by adding instructions to a file called `.Rprofile`. Normally, when R is launched, it looks for `.Rprofile` and follows any instructions the file contains, including sourcing `renv/activate.R`.

If R is launched from `Rscript` or `littler`, however, it does *not* look for an `.Rprofile` file. Since it ignores `.Rprofile`, it does not run `renv/activate.R`, and `renv` does not work.


# `command::use_renv()` fixes the problem

The problem of `Rscript` and `littler` bypassing `renv/activate.R` can be fixed by including the line `command::use_renv()` at the top of each R script in the project, as in,
```{r, eval = FALSE}
command::use_renv()

suppressPackageStartupMessages({
  library(dplyr)
  library(command)
})

cmd_assign(.data = "data/cleaned.rds",
           n_iter = 2000L,
           .out = "out/model.rds")

# ... rest of script ...
```

Function `use_renv()` looks for `renv` infrastructure and, if it finds it, sources `renv/activate.R`. The sourcing occurs later than it would in a normal `.Rprofile` startup, but early enough for subsequent `library()` calls to see the project library.

If `use_renv()` does not find any `renv` infrastructure, it simply returns `NULL`. Including `command::use_renv()` in projects that do not use `renv` does no harm.


# The exact format matters

The call to `command::use_renv()` must come before any calls to `library()`. If `library()` has already been called, then `renv/activate.R` does not work properly.

The prohibition against calling `library()` before `command::use_renv()` includes not calling `library(command)`. Use the `::` operator instead:
```{r, eval = FALSE}
command::use_renv()

suppressPackageStartupMessages({
  library(dplyr)
  library(command)
})

# ... rest of script ...
```
Do *not* load `command` first:
```{r, eval = FALSE}
library(command)  ## not correct!!
use_renv()        ## not correct!!

suppressPackageStartupMessages({
  library(dplyr)
  library(command)
})

# ... rest of script ...
```
