---
title: "Using command with renv"
---

```{r, include = FALSE}
library(command)

knitr::opts_chunk$set(collapse = TRUE,
                      comment = "")
```

# Introduction

Data analysis workflows built with `command` typically run each script in its own R session, using [Rscript](https://stat.ethz.ch/R-manual/R-devel/library/utils/html/Rscript.html) (or [littler](https://CRAN.R-project.org/package=littler)), often from a Makefile or shell script. That pattern works well with [renv](https://rstudio.github.io/renv/), which keeps a project-specific package library. There is a catch, though: `Rscript` does not source the project `.Rprofile`, so renv is never activated, and scripts may load the wrong packages—or fail to find packages at all.

This article explains that problem and introduces `use_renv()`, a small helper that activates a renv project at the start of a pipeline script.

# The problem

A typical renv project has a `.Rprofile` that sources `renv/activate.R` when R starts. In an interactive session, or when you start R from the project directory in the usual way, that bootstrap runs and the project library is put on the search path before any packages are loaded.

`Rscript` is different. It starts a clean session and does **not** run `.Rprofile`. So this common pipeline pattern:

```{bash, eval = FALSE}
Rscript src/model.R data/cleaned.rds out/model.rds --n_iter=2000
```

runs `src/model.R` without renv. If the script then does

```{r, eval = FALSE}
library(dplyr)
library(command)
```

those packages are resolved from the default user or system library, not from the project's renv library.

Putting activation inside `cmd_assign()` would be too late: by the time `cmd_assign()` runs, `library()` calls have usually already happened, and the library search path is already wrong.

A shell wrapper that sources `renv/activate.R` before calling `Rscript` can work, but each project tends to reinvent its own wrapper, and the wrapper does not travel with the scripts themselves.

# Why not just `source("renv/activate.R")`?

You can activate renv from inside a script with base R:

```{r, eval = FALSE}
source("renv/activate.R")
```

That avoids loading the renv package first, which is good. In practice it is easy to get wrong:

- The path is relative to the **current working directory**, not necessarily the script's location. Makefiles and ad hoc `Rscript` calls do not always use the project root as the working directory.
- Scripts in subdirectories need paths like `../renv/activate.R` or `../../renv/activate.R`, which break when the layout or the working directory changes.
- Non-renv projects need an extra `if (file.exists(...))` guard, or the script errors.
- The same boilerplate is copied into every script.

`use_renv()` is meant to replace that fragile pattern with one quiet, reusable call.

# What `use_renv()` does

`use_renv()` is a lightweight bootstrap, in the same spirit as `cmd_assign()`: few arguments, sensible defaults, and a single job.

```{r, eval = FALSE}
use_renv(project = NULL, quiet = TRUE)
```

In outline:

1. Decide where to start looking — an explicit `project` directory if you pass one; otherwise the directory of the script being run (when that can be determined), or else `getwd()`.
2. Walk up the directory tree until a renv project root is found. A directory counts as a root if it contains `renv/activate.R` or `renv.lock`.
3. If no root is found, return `NULL` (quietly, when `quiet = TRUE`). Non-renv projects are fine.
4. If renv is already active for that project, do nothing and return the root.
5. Otherwise, if `renv/activate.R` exists, `source()` it once for this R process, then return the root.

The return value is the project root path, invisibly, or `NULL` when nothing was activated.

# Recommended script pattern

Call `use_renv()` as the **first executable line**, before any `library()` calls:

```{r, eval = FALSE}
command::use_renv()

suppressPackageStartupMessages({
  library(dplyr)
  library(command)
})

cmd_assign(.data = "data/cleaned.rds",
           n_iter = 2000L,
           .out = "out/model.rds")

# ... rest of script ...
```

Using `command::use_renv()` (with the `command::` prefix) means the function can run before `library(command)`. After activation, later `library()` calls see the project library.

The same pattern works whether you run the script with `Rscript` from the project root, from a subdirectory (when discovery can walk up to the root), or from a Makefile whose working directory is the project root.

# Finding `command` before renv is active

There is an unavoidable ordering issue: to call `command::use_renv()`, R must already be able to find the `command` package—typically because it is installed in your user or system library (for example via `install.packages("command")` from CRAN).

That is the usual setup for a CRAN package used across many projects. Once `use_renv()` has sourced `renv/activate.R`, the project library is prepended to `.libPaths()`, and project-specific packages load from there as intended.

If `command` existed *only* inside a project's renv library, and nowhere else, then `command::use_renv()` could not run until renv was already active. In normal use that does not arise.

# What `use_renv()` does not do

To keep the helper small, `use_renv()` does **not**:

- call `renv::restore()` or install packages
- replace `.Rprofile`, `renv::run()`, or a Makefile-level wrapper for every possible workflow
- activate other environment managers (conda, poetry, and so on)
- run automatically inside `cmd_assign()`

Those remain out of scope. The goal is one reliable line at the top of a pipeline script.

# Summary

- `Rscript` skips `.Rprofile`, so renv is not activated for typical `command` pipeline scripts.
- Hard-coding `source("renv/activate.R")` is brittle across working directories and project layouts.
- `use_renv()` discovers a renv project (when present), sources `renv/activate.R` if needed, and otherwise does nothing.
- Put `command::use_renv()` first in each pipeline script, then load packages, then call `cmd_assign()`.
